### 2017-04-21 paper.tex
Initial commit

\documentclass[10pt]{llncs}

\usepackage{cite}
    \usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{times}
%\usepackage{tabularx}
\usepackage{amsmath}
\usepackage{amsfonts}
%\usepackage{url}
\usepackage{caption}
\usepackage{multirow}
\usepackage{paralist}
\usepackage{listings}

\usepackage{color}
\definecolor{RED}{rgb}{1,0,0}
\definecolor{BLUE}{rgb}{0,0,1}
\newcommand{\FIXME}[1]{\textbf{\color{BLUE}{FIXME: #1}}}
\newcommand{\sref}[1]{Section~\ref{#1}}
\newcommand{\lref}[1]{Listing~\ref{#1}}
\newcommand{\fig}[1]{Figure~\ref{#1}}

% suppress  single floating lines on top (widow) and bottom (club)
%  10000 is infinity
%  tradeoff: possible underful vboxes
\clubpenalty=10000
\widowpenalty=10000

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
\lstset{language=[11]C++,basicstyle=\scriptsize,captionpos=b}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{From Big Data to Big Displays\\
High-Performance Visualization at Blue Brain}

%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%
%
%\IEEEcompsocitemizethanks is a special \thanks that produces the bulleted
% lists the Computer Society journals use for "first footnote" author
% affiliations. Use \IEEEcompsocthanksitem which works much like \item
% for each affiliation group. When not in compsoc mode,
% \IEEEcompsocitemizethanks becomes like \thanks and
% \IEEEcompsocthanksitem becomes a line break with idention. This
% facilitates dual compilation, although admittedly the differences in the
% desired content of \author between the different types of papers makes a
% one-size-fits-all approach a daunting prospect. For instance, compsoc
% journal papers have the author affiliations above the "Manuscript
% received ..."  text while in non-compsoc journals this is reversed. Sigh.

%% Author and Affiliation (multiple authors with multiple affiliations)
\author{Stefan~Eilemann, et.al.}
\institute{Blue Brain Project, Ecole Polytechnique Federale de Lausanne}

% make the title area
\maketitle

\begin{abstract}

  The Blue Brain project has been investing in high-performance visualization
  (HPV) to complement its HPC strategy since its inception in 2007. In 2011,
  this strategy has been accelerated to develop innovative visualization
  solutions through through increased funding and strategic partnerships with
  other research institutions.

  We present the key elements of this HPV ecosystem, which integrates C++
  visualization applications with novel collaborative display systems. We
  motivate how our strategy of transforming high-performance visualization
  engines into services enables a variety of use cases, not only for the
  integration with high-fidelity displays, but also to build service oriented
  architectures, to link into web applications and to provide remote services to
  python applications.

\end{abstract}

\section{Collaborative Display Systems}

Collaborative display systems are the integration point of the Blue Brain
visualization infrastructure. They are the evolution of existing visualization
systems. Compared to the current single-user or single-presenter systems,
collaborative display systems enable real team work through a combination of
size, resolution and user friendly implementation. Compared to immersive
visualization systems like the CAVE, they provide a simple to use environment
for high-fidelity visualization. For all use cases, the increased display size
and resolution allows improved data exploration for 2D and 3D content.

\subsection{Tiled Multitouch Display Walls}

The core of the Blue Brain visualization infrastructure are multiple
high-resolution tiled display walls driven by our Tide software~\cite{tide}. All
walls are equipped with a multitouch user interface and can be remote controlled
from any web browser. The walls are build using thin-bezel, 55 inch, Full HD LCD
panels with a hardened glass sheet. We use $4\times 3$ and $3\times 3$
configurations for a total of 24 and 18 Megapixel resolution, respectively. The
display size of over five meter diagonally (four meter for $3\times 3$) allows
team-size collaboration (up to ten people) or project-wide presentations (up to
a hundred people). \fig{fTDW} shows one wall during a project-wide presentation
with multiple interactive applications.

\begin{figure}[h!t]
  \includegraphics[width=\textwidth]{images/tdw}
  \caption{\label{fTDW}Blue Brain $4\times 3$ tiled display wall}
\end{figure}

In the last years we created a production-ready setup of monoscopic tiled
display walls. On one hand, while there is substantial research on tiled display
wall software, we found that most solutions where not ready to be used in
$24\times 7$ unattained environment. On the other hand, the technology has been
commoditized to make these type of installations affordable to medium-sized
institutions which allowed us to build the software integration for a reasonable
startup cost. The multitouch user interface implements a low entry barrier for
new users, which is a unique capability of our solution.

\subsection{OpenDeck}

OpenDeck is our next-generation visualization system, aiming to integrate the
success of tiled display walls with a seamless transition to fully immersive
environments. We are currently in the process of installing the system which
consists of a semi-cylindrical back-projection screen with a 41 Megapixel usable
resolution on a 36 $m^2$ surface (\fig{fOpenDeck}). Like the display walls, it
is equipped with multitouch capabilities which makes it usable as a monoscopic
collaboration system from the first day of installations. Unlike tiled display
walls, it is active stereo capable and is equipped with a 3D tracking system for
immersive rendering. For increased immersion, a lower resolution front
projection system fills in the floor area. OpenDeck will allow us to run our
applications based on Equalizer~\cite{EMP:09} once the system is installed.

\begin{figure}[h!t]
  \includegraphics[width=\textwidth]{images/opendeck}
  \caption{\label{fOpenDeck}OpenDeck concept rendering}
\end{figure}

The OpenDeck will provide a unique environment for the research and development
of new visualization techniques. The main area of research will be mixed reality
systems, such as touch interface implementation in an immersive environment, and
the transition from 2D content to full immersion. The OpenDeck immersive
infrastructure with multitouch will open a set of questions along immersive
touch user interfaces, transitions and mixing of monoscopic to immersive usage,
the combination of tracked and touch devices, multi-user immersion, latency for
remote immersive rendering as well as multi-site collaboration.

\subsection{Tide}

Tide (Tiled Interactive Display Environment) is the software driving the Blue
Brain tiled display walls and OpenDeck. It provides multi-window, multi-user
touch interaction on large surfaces --- think of a giant collaborative
wall-mounted tablet. Tide is a distributed application that can run on multiple
machines to power display walls or projection systems of any size. Its user
interface is designed to offer an intuitive experience on touch walls. It works
just as well on non touch-capable installations by using its web interface from
any web browser. \fig{fTDW} shows Tide on a $4\times 3$ display wall and
\fig{fTideWeb} shows the Tide web interface in a browser.

\begin{figure}[h!t]
  \includegraphics[width=\textwidth]{images/tideweb}
  \caption{\label{fTideWeb}Tide web interface}
\end{figure}

Compared to other solutions~\cite{Sage,Sage2,DisplayCluster}, the development of
Tide was driven by the need of a stable and easy to use implementation. Tide
supports three types of content: files (high-resolution images, movies, pdfs),
built-in applications (web browser, whiteboard) and remote applications using
the Deflect library (DesktopStreamer, Equalizer-based applications, Brayns).
The multitouch user interface can handle multiple users manipulating different
windows and their content.

\subsection{Deflect}

Deflect is the client library for Tide. It provides an API for pixel streaming
to Tide and for receiving events from Tide. The pixel streaming allows
synchronized parallel streaming from a parallel rendering application as well
as monoscopic and stereoscopic streams. Various events allow the application to
react to multi-touch input from the wall.

Deflect is integrated into the Equalizer parallel rendering
framework~\cite{EMP:09}, enabling transparent usage of Equalizer applications on
Tide walls. Furthermore, the DesktopStreamer application mirrors the desktop of
other machines onto a wall window and allows interaction with the remote
desktop. Other rendering applications, such as our interactive raytracing
engine Brayns~\cite{brayns} are easily integrated with Deflect and Tide.

\section{Messaging/Services}

\section{Rendering Applications}
\subsection{Interactive Raytracing}

\subsection{Out-of-Core Volume Rendering}

Real-time rendering of volumetric data suffers from the curse of
dimensionality. The data size of regular grids increases cubically as the number
of dimensions increase. To render such data interactively, out-of-core (OOC)
storage using multi-resolution algorithms are commonly employed.

Livre is an interactive volume rendering engine available under a permissive
open source license. Our main contribution are: an state-of-the-art
implementation of an octree-based level-of-detail selection, a task parallel
rendering pipeline, a multi-GPU based parallel rendering engine, and an easily
extensible renderer through the use of plugin data sources. Furthermore we
propose to create volume representations of different input data using
on-the-fly, interactive transformations of the data, skipping costly
preprocessing steps.

Our system brings together state-of-the art algorithms to create a volume
rendering engine capable of handling extremely high-resolution volumes using a
high degree of parallelism, both on a single system as well as in a distributed
cluster. We employ a GPU-based ray casting algorithm to compute the radiance
absorption of the given volumetric data. The computation is executed per pixel
on the pixel shader hardware of the GPU.

In our out-of-core data access layer, multi-resolution data is represented as an
octree data structure. This representation accelerates the selection of the
proper level-of-detail and to track the status of the LODs (in CPU memory, in
GPU memory, not loaded). While rendering, view-based LOD selection is performed
using the on the screen-space-error (SSE)~\cite{guthe2004} technique. According
to the requested maximum SSE, the selected LODs represent the given volume the
best on the screen.

The creation of volume bricks, their upload to the GPU and the rendering are
executed in separate tasks. These tasks run asynchronously, that is, they do not
block each other. This is particularly important for the render thread, which
has to react to user input as quickly as possible to reduce input latency, and
renders at interactive frame rates. The data loader thread can take seconds to
generate the brick, for example in the field voxelization data source when
sampling millions of events.

Livre uses a plugin mechanism to access the volume data. In this mechanism, data
sources are implemented as shared libraries and are loaded on application
startup based on the URI of the input data. Loading the shared libraries on
runtime allows to add third-party data loaders to an existing Livre application
by simply placing them into the shared library directory. Data sources have to
provide the requested volume bricks, that is, there is no defined file format or
even requirement to read the input data from a file system. This flexibility of
the plugin approach lead to novel volume rendering use cases, where volume
representations are created on the fly from different input data sets.

%----------------------------------------------------------------------
\section{Discussion and Conclusion}
%----------------------------------------------------------------------
\label{sec:conclusions}

%----------------------------------------------------------------------
\section*{Acknowledgments}
%----------------------------------------------------------------------
This publication was supported by the Blue Brain Project (BBP), the Swiss
National Science Foundation under Grant 200020-129525, the King Abdullah
University of Science and Technology (KAUST) through the KAUST-EPFL alliance for
Neuro-Inspired High Performance Computing, the Spanish Ministry of Science and
Innovation under grant (TIN2010-21289-C02-01/02), the Cajal Blue Brain Project,
Hasler and HBP.

\bibliographystyle{abbrv}
\bibliography{references/references}

% that's all folks
\end{document}

### 2017-04-21 paper.tex
Draft messaging section

@@
 \begin{abstract}
-
   The Blue Brain project has been investing in high-performance visualization
@@
   python applications.
-
 \end{abstract}
 
-\section{Collaborative Display Systems}
+\section{Collaborative Displays}
 
@@
 \begin{figure}[h!t]
-  \includegraphics[width=\textwidth]{images/tdw}
+  \includegraphics[width=\columnwidth]{images/tdw}
   \caption{\label{fTDW}Blue Brain $4\times 3$ tiled display wall}
@@
 \begin{figure}[h!t]
-  \includegraphics[width=\textwidth]{images/opendeck}
+  \includegraphics[width=\columnwidth]{images/opendeck}
   \caption{\label{fOpenDeck}OpenDeck concept rendering}
@@
 \begin{figure}[h!t]
-  \includegraphics[width=\textwidth]{images/tideweb}
+  \includegraphics[width=\columnwidth]{images/tideweb}
   \caption{\label{fTideWeb}Tide web interface}
@@
 
-\section{Messaging/Services}
+\section{C++ Messaging and Services}
+
+All Blue Brain applications integrate a messaging framework which allows them
+to be used as services in a variety of use cases. For example, the Tide web
+server providing the user interface shown in \fig{fTideWeb}, is based on this
+messaging solution. Other use cases are remote python APIs, JavaScript user
+interfaces and service architectures combining multiple visualization
+applications with data providers such as HPC simulations.
+
+The base communication layer ZeroEQ utilizes ZeroMQ as the transport layer, the
+ZeroConf protocol for discovery, and our novel ZeroBuf serialization library for
+high-performance messaging. A fully integrated HTTP server provides a bridge to
+JavaScript, Python and similar environments by implementing REST APIs with JSON
+payload. \fig{fUML} shows a class diagram of our messaging solution.
+
+\begin{figure*}[ht]\center
+  \includegraphics[width=\columnwidth]{images/ZeroMS}
+  \caption{\label{fUML}UML Diagram of the main messaging classes}
+\end{figure*}
+
+\subsection{ZeroEQ}
+
+ZeroEQ is the base messaging library, wrapping up existing technologies into an
+API which is convenient to use and easy to integrate into C++ code. It provides
+two messaging services: publish-subscribe and HTTP. For binary and JSON
+encodings it relies on a simple \textsf{Serializable} interface, for which
+ZeroBuf provides a sample implementation. To facilitate the simple use case of
+linking a few applications, ZeroEQ uses the zeroconf protocol to discover and
+connect to related applications. For more complex scenarios, explicit
+connections are supported.
+
+\subsubsection{Publish-Subscribe}
+
+The publish-subscribe service is implemented in a \textsf{Publisher} and
+\textsf{Subscriber} class. It provides event-based messaging, based on a 128-bit
+message type with arbitrary payload. The message type is used for message
+subscription, filtering and routing. The payload is expected to be uniquely
+identified by the message type, that is, all applications agree for the decoding
+and semantics of any given message type. The ZeroBuf section outlines how this
+is implemented. The underlying transport uses ZeroMQ pub-sub sockets.
+
+The pub-sub service is stateless, that is, applications have no expectation of
+when messages are received or who receives published messages. This
+communication pattern naturally leads to robust services, since there is no
+possiblity for deadlocks or undefined behaviour. The pub-sub API is provided in
+two flavors: a simple $pointer \& size$ memory buffer, and a higher level
+object-based abstraction. The object-based API is syntactic sugar for the
+low-level API, and allows automatic publish and update of objects with a few
+lines of code. It uses the \textsf{toBinary()} and \textsf{fromBinary()} methods
+of the \textsf{Serializable} interface to call the low-level API.
+
+\begin{lstlisting}[float, caption=Publish-Subscribe Example, label=lPubSub]
+zerobuf::render::Camera camera;
+zeroeq::Publisher publisher;
+zeroeq::Subscriber subscriber;
+
+subscriber.subscribe( camera );
+
+while( rendering )
+{
+    subscriber.receive( 0 /*ms, poll*/ );
+    updateCamera( camera );
+    publisher.publish( camera );
+    renderFrame( camera );
+}
+\end{lstlisting}
+
+The example in \lref{lPubSub} shows the integration of camera
+synchronization in a visualization application. This example relies on the
+builtin zeroconf protocol to connect application instances. Subscribers only
+subscribe to events from publishers within the same session. The default
+session name is the user name, and can be customized using an environment
+variable or non-default constructor. Similarly, the subscriber can subscribe by
+session or address. \lref{lPubSubExp} illustrates an explicitly addressed
+subscription. Notice that the subscriber uses the publisher URI, which will
+contain the concrete port chosen for the publisher.
+
+\begin{lstlisting}[float, caption=Explicit Addressing,label=lPubSubExp]
+zeroeq::URI uri( "tcp://localhost" );
+zeroeq::Publisher publisher( uri, zeroeq::NULL_SESSION ); // deactivate zeroconf
+zeroeq::Subscriber subscriber( publisher.getURI( )); // use concrete port
+\end{lstlisting}
+
+Subscribers are derived from a \textsf(Receiver) base class, which is shared
+with the http server. All receivers can share their \textsf{receive()}
+operation at construction time, that is, the blocking receive operation applies
+to all receivers in the shared group. \lref{lSubShare} shows an example of
+selectively receiving different updates on different input sockets.
+
+\begin{lstlisting}[float,caption=Subscriber Sharing, label=lSubShare]
+zeroeq::Subscriber local( zeroeq::URI( "localhost:29387" ));
+zeroeq::Subscriber global( local );
+
+local.subscribe( colorMap );
+global.subscribe( camera );
+
+while( true )
+    local.receive(); // updates colorMap and camera
+\end{lstlisting}
+
+\subsubsection{HTTP Server}
+
+The http server is built using cppnetlib \cite{cppnetlib} for the transport and
+http protocol handling. It supports all standard http verbs (GET, POST, PUT,
+PATCH, DELETE). It is a \textsf{zeroeq::Receiver}, that is, it can share its
+\textsf{receive()} update operation with other subscribers and http servers.
+Unlike a subscriber, the http server follows the HTTP request-reply semantics,
+that is, a request received by a server has to be followed directly by its
+reply. future
+
+vocabulary
 

### 2017-05-10 paper.tex
http server, python API

@@
 
-The http server is built using cppnetlib \cite{cppnetlib} for the transport and
+The http server is built using cppnetlib~\cite{cppnetlib} for the transport and
 http protocol handling. It supports all standard http verbs (GET, POST, PUT,
@@
 that is, a request received by a server has to be followed directly by its
-reply. future
+reply. To allow asynchronous request processing, the return value from the
+request handler is a \textsf{std::future} which is retrieved from an internal
+thread, thus allowing the application to continue operations.
 
-vocabulary
+The http server is introspectable, it allows querying the available endpoints
+(objects) and the JSON schema~\cite{jsonschema} for each endpoint.
+\lref{lhttpReg} shows an excerpt of the Tide registry, and \lref{lhttpSchema} an
+excerpt of the schema for one of the exposed objects. This REST API is used by
+the Tide web interface from Javascript and to generate remote python APIs.
+
+\noindent\begin{minipage}[b][][b]{.48\textwidth}
+\begin{lstlisting}[caption=HTTP Server Registry, label=lhttpReg]
+> GET /registry HTTP/1.0
+{
+[...]
+   "tide/open" : [ "PUT" ],
+   "tide/options" : [ "GET", "PUT" ],
+   "tide/resize-window" : [ "PUT" ],
+[...]
+}
+\end{lstlisting}
+\end{minipage}\hfill
+\begin{minipage}[b][][b]{.48\textwidth}
+\begin{lstlisting}[caption=Object JSON Schema, label=lhttpSchema]
+> GET /tide/options/schema HTTP/1.0
+{
+[...]
+    "properties": {
+        "alphaBlending": {
+            "type": "boolean"
+        },
+[...]
+}
+\end{lstlisting}
+\end{minipage}
+
+\subsection{Remote Python API}
+
+The remote python API provides easy to use access to remote applications using
+the http server. It integrates two features: generic code generation for the
+REST API exposed by the application, and automatic resource allocation and
+application launch.
+
+The generic code generation is implemented in a pure python module, which has
+no dependency to the interfaced C++ application. It queries the http server and
+generates a python API for all exposed objects. This API can then be
+conveniently used in python to remote control the application.
+
+Access to the application is established either through an explicit connection
+of a pre-launched application, or via a resource allocator. The allocator hides
+the details of allocating a resource, e.g. using a scheduling system like
+slurm, launching and connecting to the launched application from the python
+programmer.
 
 \section{Rendering Applications}
+
+The rendering applications form the backbone of our ecosystem.
+
 \subsection{Interactive Raytracing}
+
+\subsection{OpenGL Parallel Rendering}
 

### 2017-05-10 paper.tex
Brayns

@@
 
+\FIXME{teaser gallery}
+
 \section{Rendering Applications}
 
-The rendering applications form the backbone of our ecosystem.
+The rendering applications form the backbone of our ecosystem. They cover a
+wide range of established rendering algorithms to serve a broad set of use
+cases for visual debugging, scientific illustrations and outreach.
 
-\subsection{Interactive Raytracing}
+\subsection{Brayns: Interactive Raytracing}
 
-\subsection{OpenGL Parallel Rendering}
+Advances in computer hardware and software have brought raytracing to the point
+where it replaces classical rasterization for virtually all use cases in
+scientific visualization. On one hand, local illumination for typical OpenGL
+data sets (up to hundred million triangles) can be done at similar framerates
+to OpenGL~\cite{ospray}. For small datasets, OpenGL performs better, but for
+larger data sets raytracing outperforms OpenGL. This is due to a better
+scalability with respect to the data set size ($O(\log{N})$ vs $O(n)$).
+Furthermore, CPU-based raytracing allows the rendering of larger data sets
+without any level-of-detail algorithms due to the larger memory size. Last, but
+not least, advanced rendering algorithms such as shadowing, reflections and
+global illuminations are significantly easier to implement in a raytracer.
 
-\subsection{Out-of-Core Volume Rendering}
+It is for these reasons that interactive raytracing is the future technology for
+interactive and offline rendering at Blue Brain. We have developed a first open
+source implementation of a visualization engine with different backends:
+OSPRay~\cite{ospray} for CPU-based raytracing and OptiX~\cite{optix} for
+GPU-based raytracing. These application called Brayns can load and visualize
+our data sets in a variety of modes and integrates with our messaging solution
+to be accessible from Python.
+
+\FIXME{teaser gallery}
+
+\subsection{RTNeuron: OpenGL Parallel Rendering}
+
+\subsection{Livre: Out-of-Core Volume Rendering}
 

### 2017-05-10 paper.tex
Tweaks

@@
 %% Author and Affiliation (multiple authors with multiple affiliations)
-\author{Stefan~Eilemann, et.al.}
+\author{Stefan~Eilemann, et.al.\thanks{\textit{firstname.lastname}@epfl.ch}}
 \institute{Blue Brain Project, Ecole Polytechnique Federale de Lausanne}
@@
 not least, advanced rendering algorithms such as shadowing, reflections and
-global illuminations are significantly easier to implement in a raytracer.
+global illuminations are significantly easier to implement in a raytracer. The
+only area where rasterization provides a benefit is for rendering at very high
+frame rates, needed for example for immersive visualization.
 
@@
 our data sets in a variety of modes and integrates with our messaging solution
-to be accessible from Python.
+to be accessible from Python. We are currently integrating the key algorithms
+from Livre for out-of-core volume rendering in Brayns.
 
@@
 Livre is an interactive volume rendering engine available under a permissive
-open source license. Our main contribution are: an state-of-the-art
+open source license. Our main contributions are: an state-of-the-art
 implementation of an octree-based level-of-detail selection, a task parallel
@@
 
+\FIXME{teaser gallery}
+
 %----------------------------------------------------------------------
@@
 \label{sec:conclusions}
+
+Brayns will become the integrative visualizer
 

### 2017-05-15 paper.tex
Jupyter image
Review and ZeroBuf

@@
 \usepackage{times}
-%\usepackage{tabularx}
 \usepackage{amsmath}
 \usepackage{amsfonts}
-%\usepackage{url}
 \usepackage{caption}
@@
 \usepackage{listings}
+\usepackage{wrapfig}
 
@@
 \widowpenalty=10000
+
+\setlength{\columnsep}{2em}% wrapfig margin, among others
 
@@
 
-\section{Collaborative Displays}
+\section{High-Fidelity Displays}
 
-Collaborative display systems are the integration point of the Blue Brain
+High-fidelity display systems are the integration point of the Blue Brain
 visualization infrastructure. They are the evolution of existing visualization
-systems. Compared to the current single-user or single-presenter systems,
-collaborative display systems enable real team work through a combination of
-size, resolution and user friendly implementation. Compared to immersive
-visualization systems like the CAVE, they provide a simple to use environment
-for high-fidelity visualization. For all use cases, the increased display size
-and resolution allows improved data exploration for 2D and 3D content.
+systems, enabling higher resolution, immersion and collaboration. Compared to
+the current single-user or single-presenter systems, collaborative display
+systems allow real team work through a combination of size, resolution and user
+friendly implementation. Compared to immersive visualization systems like the
+CAVE, they provide a simple to use environment for high-fidelity visualization.
+For all use cases, the increased display size and resolution allows improved
+data exploration for 2D and 3D content.
 
@@
 a hundred people). \fig{fTDW} shows one wall during a project-wide presentation
-with multiple interactive applications.
+with multiple interactive applications running concurrently on the wall.
 
@@
 wall software, we found that most solutions where not ready to be used in
-$24\times 7$ unattained environment. On the other hand, the technology has been
-commoditized to make these type of installations affordable to medium-sized
-institutions which allowed us to build the software integration for a reasonable
-startup cost. The multitouch user interface implements a low entry barrier for
-new users, which is a unique capability of our solution.
+$24\times 7$ unattained environment. For this reason, we started with the TACC
+DisplayCluster open source software~\cite{DisplayCluster}, which we
+incrementally refactored and improved to the current TIDE implementation.
+
+On the other hand, the technology has been commoditized to make these type of
+installations affordable to medium-sized institutions which allowed us to build
+the software integration for a reasonable startup cost. We have focused on the
+multitouch user interface, which implements a low entry barrier for new users,
+a unique capability of our solution.
 
@@
 environments. We are currently in the process of installing the system which
-consists of a semi-cylindrical back-projection screen with a 41 Megapixel usable
+consists of a semi-cylindrical back-projection screen with 41 Megapixel usable
 resolution on a 36 $m^2$ surface (\fig{fOpenDeck}). Like the display walls, it
 is equipped with multitouch capabilities which makes it usable as a monoscopic
-collaboration system from the first day of installations. Unlike tiled display
-walls, it is active stereo capable and is equipped with a 3D tracking system for
+collaboration system from the first day of installation. Unlike tiled display
+walls, it is active stereo capable and equipped with a 3D tracking system for
 immersive rendering. For increased immersion, a lower resolution front
@@
 The OpenDeck will provide a unique environment for the research and development
-of new visualization techniques. The main area of research will be mixed reality
-systems, such as touch interface implementation in an immersive environment, and
-the transition from 2D content to full immersion. The OpenDeck immersive
-infrastructure with multitouch will open a set of questions along immersive
-touch user interfaces, transitions and mixing of monoscopic to immersive usage,
-the combination of tracked and touch devices, multi-user immersion, latency for
-remote immersive rendering as well as multi-site collaboration.
+of new visualization techniques. The OpenDeck immersive infrastructure with
+multitouch will open a set of questions along immersive touch user interfaces,
+transitions and mixing of monoscopic to immersive usage, the combination of
+tracked and touch devices, multi-user immersion, latency for remote immersive
+rendering as well as multi-site collaboration.
 
@@
 to Tide and for receiving events from Tide. The pixel streaming allows
-synchronized parallel streaming from a parallel rendering application as well
-as monoscopic and stereoscopic streams. Various events allow the application to
+synchronized parallel streaming from a parallel rendering application as well as
+monoscopic and stereoscopic streams. Various events allow the application to
 react to multi-touch input from the wall.
@@
 
-All Blue Brain applications integrate a messaging framework which allows them
+\begin{wrapfigure}{r}{.382\textwidth}\center\vspace{-6ex}
+  \includegraphics[width=.382\textwidth]{images/ZeroMS}
+  {\caption{\label{fZero}Messaging libraries and use cases}}\vspace{-2ex}
+\end{wrapfigure}
+
+All Blue Brain applications integrate a messaging libraries which allow them
 to be used as services in a variety of use cases. For example, the Tide web
@@
 interfaces and service architectures combining multiple visualization
-applications with data providers such as HPC simulations.
+applications with data providers such as HPC simulations (\fig{fZero}).
 
@@
 
-\begin{figure*}[ht]\center
-  \includegraphics[width=\columnwidth]{images/ZeroMS}
+\begin{figure}[ht]\center
+  \includegraphics[width=\columnwidth]{images/ZeroMSUML}
   \caption{\label{fUML}UML Diagram of the main messaging classes}
-\end{figure*}
+\end{figure}
 
@@
 identified by the message type, that is, all applications agree for the decoding
-and semantics of any given message type. The ZeroBuf section outlines how this
-is implemented. The underlying transport uses ZeroMQ pub-sub sockets.
+and semantics of any given message type. ZeroBuf provides a sample
+implementation for this. The underlying transport uses ZeroMQ pub-sub sockets.
 
@@
 
-The http server is introspectable, it allows querying the available endpoints
-(objects) and the JSON schema~\cite{jsonschema} for each endpoint.
-\lref{lhttpReg} shows an excerpt of the Tide registry, and \lref{lhttpSchema} an
-excerpt of the schema for one of the exposed objects. This REST API is used by
-the Tide web interface from Javascript and to generate remote python APIs.
+The data served by the http server is introspectable, it allows querying the
+available endpoints (objects) and the JSON schema~\cite{jsonschema} for each
+endpoint. \lref{lhttpReg} shows an excerpt of the Tide registry, and
+\lref{lhttpSchema} an excerpt of the schema for one of the exposed objects. This
+REST API is used by the Tide web interface from Javascript and to generate
+remote python APIs.
 
@@
 
+\subsection{ZeroBuf}
+
+ZeroBuf is a sample implementation of serialization for ZeroEQ. Based on a
+grammar closely related to Flatbuffers schemas~\cite{flatbuffers}, it generates
+C++ classes with random \textsf{set/get} access. All data is stored internally
+in one continuous memory buffer, which can be used for the ZeroEQ binary
+serialization without any copy. The conversion to and from a JSON representation
+involves a copy using a \textsf{std::string}. ZeroBuf can store:
+
+\begin{compactitem}
+\item (u)int[8,16,32,64,128]\_t, float, double and string members
+\item fixed size arrays and dynamic vectors of static-sized elements (intrinsic members or composite types)
+\item static and dynamic sub-classes (composite types of the above)
+\end{compactitem}
+
+\begin{figure}[ht]\center
+  \includegraphics[width=\columnwidth]{images/ZeroBufStatic}\\\vspace{4ex}
+  \includegraphics[width=\columnwidth]{images/ZeroBufDynamic}
+  \caption{\label{fZeruBuf}ZeroBuf Examples for static (top) and dynamic (bottom) sized objects}
+\end{figure}
+
 \subsection{Remote Python API}
@@
 
-\FIXME{teaser gallery}
+\fig{fJupyter} shows an example session of using this Python API from a Jupyter
+notebook, allocating and launching a Brayns instance, setting relevant
+rendering parameters and retrieving an image.
+
+\begin{figure}[ht]\center
+  \includegraphics[width=\columnwidth]{images/jupyter}
+  \caption{\label{fJupyter}Example Jupyter notebook session using Brayns}
+\end{figure}
 

### 2017-05-29 paper.tex
Motivation

@@
 % correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
+\hyphenation{op-tical net-works semi-conduc-tor Zero-Buf}
 
@@
 
+\section{Motivation}
+
+The Blue Brain Project (BBP) uses simulation-based research to analyze and reverse
+engineer cortical neuron circuits. The simulated models go beyond using detailed
+models of individual neurons or large-scale network models of simplified
+neurons, they model in the order of hundreds of thousands of detailed neurons in
+a fully connected circuit. The project generates a multitude of data for the
+model building and during the simulation of these models. This data ranges from
+terabyte-sized image stacks for data extraction to detailed in-silico circuit
+models of large geometric complexity and terabyte-size simulation reports.
+
+Visualization supports the BBP along all parts of this project to understand and
+debug model data, building and simulation algorithms as well as validating and
+discovering new insight from the in-silico experiments. Our strategy to support
+this mission is based on components linked through network protocols:
+High-fidelity display systems to see more detail in complex data, a set of
+standard rendering engines (rasterization, out-of-core volume rendering,
+interactive raytracing), and decoupled, light-weight applications using remote
+these components. In the following we will present these components along with a
+few use cases.
+
 \section{High-Fidelity Displays}
@@
 visualization infrastructure. They are the evolution of existing visualization
-systems, enabling higher resolution, immersion and collaboration. Compared to
-the current single-user or single-presenter systems, collaborative display
+systems, enabling higher resolution, immersion and team collaboration. Compared
+to current single-user or single-presenter systems, collaborative display
 systems allow real team work through a combination of size, resolution and user
@@
 For all use cases, the increased display size and resolution allows improved
-data exploration for 2D and 3D content.
+data exploration for 2D and 3D content, facilitating large data exploration.
 
@@
 
-In the last years we created a production-ready setup of monoscopic tiled
+In the last years we implemented a production-ready setup of monoscopic tiled
 display walls. On one hand, while there is substantial research on tiled display
-wall software, we found that most solutions where not ready to be used in
-$24\times 7$ unattained environment. For this reason, we started with the TACC
-DisplayCluster open source software~\cite{DisplayCluster}, which we
-incrementally refactored and improved to the current TIDE implementation.
+wall software~\cite{Sage, Sage2, Omegalib, DK:11, JLMV:06, DisplayCluster}, we
+found that most solutions where not ready to be used in a $24\times 7$
+unattained environment. For this reason, we started with the TACC DisplayCluster
+open source software~\cite{DisplayCluster}, which we incrementally refactored
+and improved to the current TIDE implementation.
 
@@
   \includegraphics[width=.382\textwidth]{images/ZeroMS}
-  {\caption{\label{fZero}Messaging libraries and use cases}}\vspace{-2ex}
+  {\caption{\label{fZero}Messaging enables new use cases}}\vspace{-2ex}
 \end{wrapfigure}
@@
 possiblity for deadlocks or undefined behaviour. The pub-sub API is provided in
-two flavors: a simple $pointer \& size$ memory buffer, and a higher level
+two flavors: a simple $pointer\ \&\ size$ memory buffer, and a higher level
 object-based abstraction. The object-based API is syntactic sugar for the
@@
 \item (u)int[8,16,32,64,128]\_t, float, double and string members
-\item fixed size arrays and dynamic vectors of static-sized elements (intrinsic members or composite types)
+\item fixed size arrays and dynamic vectors of static-sized elements (intrinsic
+members or composite types)
 \item static and dynamic sub-classes (composite types of the above)
@@
   \includegraphics[width=\columnwidth]{images/ZeroBufDynamic}
-  \caption{\label{fZeruBuf}ZeroBuf Examples for static (top) and dynamic (bottom) sized objects}
+  \caption{\label{fZeroBuf}ZeroBuf Examples for static (top) and dynamic
+(bottom) sized objects}
 \end{figure}
+
+\fig{fZeroBuf} shows two simple ZeroBuf schemas together with example usage of
+the generated code in C++ and their memory layout. The static example shows
+nested ZeroBuf classes for the camera used in \lref{lPubSub}, and the dynamic
+example shows how raw data access is used to prepare a JPEG image for
+publishing.
 
@@
 \fig{fJupyter} shows an example session of using this Python API from a Jupyter
-notebook, allocating and launching a Brayns instance, setting relevant
-rendering parameters and retrieving an image.
+notebook, allocating and launching a Brayns instance, setting relevant rendering
+parameters and retrieving an image. Note that the whole notebook runs in a
+light-weight VM with no GPU and interacts with a Brayns instance launched on a
+bare-metal visualization cluster node.
 
@@
 wide range of established rendering algorithms to serve a broad set of use
-cases for visual debugging, scientific illustrations and outreach.
+cases for visual debugging, scientific illustrations and communication.
 
@@
 
+Messaging enables surprising new use cases
+
+Tiled display walls are cheap and ready for new ways of working
+
 %----------------------------------------------------------------------

### 2017-05-29 paper.tex
Cleanups

@@
 High-fidelity display systems are the integration point of the Blue Brain
-visualization infrastructure. They are the evolution of existing visualization
+visualization capabilities. They are the evolution of existing visualization
 systems, enabling higher resolution, immersion and team collaboration. Compared
@@
 friendly implementation. Compared to immersive visualization systems like the
-CAVE, they provide a simple to use environment for high-fidelity visualization.
-For all use cases, the increased display size and resolution allows improved
-data exploration for 2D and 3D content, facilitating large data exploration.
+CAVE, they provide a more approachable environment for high-fidelity
+visualization. For all use cases, the increased display size and resolution
+allows better data exploration for 2D and 3D content, facilitating large data
+exploration.
 
@@
 a hundred people). \fig{fTDW} shows one wall during a project-wide presentation
-with multiple interactive applications running concurrently on the wall.
+with multiple interactive applications running remotely on the wall.
 
@@
 
-In the last years we implemented a production-ready setup of monoscopic tiled
-display walls. On one hand, while there is substantial research on tiled display
-wall software~\cite{Sage, Sage2, Omegalib, DK:11, JLMV:06, DisplayCluster}, we
-found that most solutions where not ready to be used in a $24\times 7$
-unattained environment. For this reason, we started with the TACC DisplayCluster
-open source software~\cite{DisplayCluster}, which we incrementally refactored
-and improved to the current TIDE implementation.
-
-On the other hand, the technology has been commoditized to make these type of
-installations affordable to medium-sized institutions which allowed us to build
-the software integration for a reasonable startup cost. We have focused on the
-multitouch user interface, which implements a low entry barrier for new users,
-a unique capability of our solution.
+While there is substantial research on tiled display wall software~\cite{Sage,
+Sage2, Omegalib, DK:11, JLMV:06, DisplayCluster}, we found that most solutions
+where not ready production use in a $24\times 7$ unattained environment. For
+this reason, we started with the TACC DisplayCluster open source
+software~\cite{DisplayCluster}, which we incrementally refactored and improved
+to the current TIDE implementation. On the other hand, the hardware has been
+commoditized to make these type of installations affordable to medium-sized
+institutions which allowed us to build the software integration for a reasonable
+startup cost. We have focused on the multitouch user interface, which implements
+a low entry barrier for new users, a unique capability of our solution.
 
@@
 success of tiled display walls with a seamless transition to fully immersive
-environments. We are currently in the process of installing the system which
+environments. We are currently in the process of installing a system which
 consists of a semi-cylindrical back-projection screen with 41 Megapixel usable
@@
 immersive rendering. For increased immersion, a lower resolution front
-projection system fills in the floor area. OpenDeck will allow us to run our
-applications based on Equalizer~\cite{EMP:09} once the system is installed.
+projection system fills in the floor area. OpenDeck will run TIDA and our
+immersive applications based on Equalizer~\cite{EMP:09} once the system is
+installed.
 
@@
 
-The OpenDeck will provide a unique environment for the research and development
-of new visualization techniques. The OpenDeck immersive infrastructure with
-multitouch will open a set of questions along immersive touch user interfaces,
-transitions and mixing of monoscopic to immersive usage, the combination of
-tracked and touch devices, multi-user immersion, latency for remote immersive
-rendering as well as multi-site collaboration.
+OpenDeck will provide a unique environment for the research and development of
+new visualization techniques. It will open a set of questions along immersive
+touch user interfaces, transitions and mixing of monoscopic to immersive usage,
+the combination of tracked and touch devices, multi-user immersion, latency for
+remote immersive rendering as well as multi-site collaboration.
 
@@
 The multitouch user interface can handle multiple users manipulating different
-windows and their content.
+windows and their content simultaneously.
 
@@
 to Tide and for receiving events from Tide. The pixel streaming allows
-synchronized parallel streaming from a parallel rendering application as well as
+synchronized parallel streaming from parallel rendering application as well as
 monoscopic and stereoscopic streams. Various events allow the application to
@@
 
+\section{Rendering Applications}
+
+The rendering applications form the backbone of our ecosystem. They cover a
+wide range of established rendering algorithms to serve a broad set of use
+cases for visual debugging, scientific illustrations and communication.
+
+\subsection{Brayns: Interactive Raytracing}
+
+Advances in computer hardware and software have brought raytracing to the point
+where it replaces classical rasterization for virtually all use cases in
+scientific visualization. On one hand, OpenGL-like local illumination for
+typical data sets used with rasterization (up to hundred million triangles) can
+be done at similar framerates to OpenGL~\cite{ospray}. For small datasets,
+OpenGL performs better, but for larger data sets raytracing outperforms OpenGL.
+This is due to a better scalability with respect to the data set size
+($O(\log{N})$ vs $O(n)$). Furthermore, CPU-based raytracing allows the rendering
+of larger data sets without any level-of-detail algorithms due to the larger
+memory size. Last, but not least, advanced rendering algorithms such as
+shadowing, reflections and global illuminations are significantly easier to
+implement in a raytracer. The only area where rasterization provides a benefit
+is for rendering at very high frame rates, needed for example for immersive
+visualization.
+
+It is for these reasons that interactive raytracing is the future technology for
+interactive and offline rendering at Blue Brain. We have developed a first open
+source implementation of a visualization engine with different backends:
+OSPRay~\cite{ospray} for CPU-based raytracing and OptiX~\cite{optix} for
+GPU-based raytracing. This application called Brayns can load and visualize our
+data sets in a variety of modes and integrates with our messaging solution to be
+accessible from Python. We are currently integrating the key algorithms from
+Livre for out-of-core volume rendering in Brayns, which will also facilitate
+mixing polygonal with volume data in a single scene.
+
+\FIXME{teaser gallery}
+
+\subsection{RTNeuron: OpenGL Parallel Rendering}
+
+\subsection{Livre: Out-of-Core Volume Rendering}
+
+Real-time rendering of volumetric data suffers from the curse of
+dimensionality. The data size of regular grids increases cubically as the number
+of dimensions increase. To render such data interactively, out-of-core (OOC)
+storage using multi-resolution algorithms are commonly employed.
+
+Livre is an interactive volume rendering engine available under a permissive
+open source license. Our main contributions are: an state-of-the-art
+implementation of an octree-based level-of-detail selection, a task parallel
+rendering pipeline, a multi-GPU based parallel rendering engine, and an easily
+extensible renderer through the use of plugin data sources. Furthermore we
+propose to create volume representations of different input data using
+on-the-fly, interactive transformations of the data, skipping costly
+preprocessing steps.
+
+Our system brings together state-of-the art algorithms to create a volume
+rendering engine capable of handling extremely high-resolution volumes using a
+high degree of parallelism, both on a single system as well as in a distributed
+cluster. We employ a GPU-based ray casting algorithm to compute the radiance
+absorption of the given volumetric data. The computation is executed per pixel
+on the pixel shader hardware of the GPU.
+
+In our out-of-core data access layer, multi-resolution data is represented as an
+octree data structure. This representation accelerates the selection of the
+proper level-of-detail and to track the status of the LODs (in CPU memory, in
+GPU memory, not loaded). While rendering, view-based LOD selection is performed
+using the on the screen-space-error (SSE)~\cite{guthe2004} technique. According
+to the requested maximum SSE, the selected LODs represent the given volume the
+best on the screen.
+
+The creation of volume bricks, their upload to the GPU and the rendering are
+executed in separate tasks. These tasks run asynchronously, that is, they do not
+block each other. This is particularly important for the render thread, which
+has to react to user input as quickly as possible to reduce input latency, and
+renders at interactive frame rates. The data loader thread can take seconds to
+generate the brick, for example in the field voxelization data source when
+sampling millions of events.
+
+Livre uses a plugin mechanism to access the volume data. In this mechanism, data
+sources are implemented as shared libraries and are loaded on application
+startup based on the URI of the input data. Loading the shared libraries on
+runtime allows to add third-party data loaders to an existing Livre application
+by simply placing them into the shared library directory. Data sources have to
+provide the requested volume bricks, that is, there is no defined file format or
+even requirement to read the input data from a file system. This flexibility of
+the plugin approach lead to novel volume rendering use cases, where volume
+representations are created on the fly from different input data sets.
+
+\FIXME{teaser gallery}
+
 \section{C++ Messaging and Services}
@@
 
-All Blue Brain applications integrate a messaging libraries which allow them
-to be used as services in a variety of use cases. For example, the Tide web
-server providing the user interface shown in \fig{fTideWeb}, is based on this
-messaging solution. Other use cases are remote python APIs, JavaScript user
-interfaces and service architectures combining multiple visualization
-applications with data providers such as HPC simulations (\fig{fZero}).
+All Blue Brain applications integrate messaging libraries which allow them to be
+used as services in a variety of use cases. For example, the Tide web server
+providing the user interface shown in \fig{fTideWeb}, is based on this messaging
+solution. Other use cases are remote python APIs, JavaScript user interfaces and
+service architectures combining multiple visualization applications with data
+providers such as HPC simulations (\fig{fZero}).
 
@@
 
-ZeroEQ is the base messaging library, wrapping up existing technologies into an
+ZeroEQ is our C++ messaging library, wrapping up existing technologies into an
 API which is convenient to use and easy to integrate into C++ code. It provides
@@
 
-\section{Rendering Applications}
-
-The rendering applications form the backbone of our ecosystem. They cover a
-wide range of established rendering algorithms to serve a broad set of use
-cases for visual debugging, scientific illustrations and communication.
-
-\subsection{Brayns: Interactive Raytracing}
-
-Advances in computer hardware and software have brought raytracing to the point
-where it replaces classical rasterization for virtually all use cases in
-scientific visualization. On one hand, local illumination for typical OpenGL
-data sets (up to hundred million triangles) can be done at similar framerates
-to OpenGL~\cite{ospray}. For small datasets, OpenGL performs better, but for
-larger data sets raytracing outperforms OpenGL. This is due to a better
-scalability with respect to the data set size ($O(\log{N})$ vs $O(n)$).
-Furthermore, CPU-based raytracing allows the rendering of larger data sets
-without any level-of-detail algorithms due to the larger memory size. Last, but
-not least, advanced rendering algorithms such as shadowing, reflections and
-global illuminations are significantly easier to implement in a raytracer. The
-only area where rasterization provides a benefit is for rendering at very high
-frame rates, needed for example for immersive visualization.
-
-It is for these reasons that interactive raytracing is the future technology for
-interactive and offline rendering at Blue Brain. We have developed a first open
-source implementation of a visualization engine with different backends:
-OSPRay~\cite{ospray} for CPU-based raytracing and OptiX~\cite{optix} for
-GPU-based raytracing. These application called Brayns can load and visualize
-our data sets in a variety of modes and integrates with our messaging solution
-to be accessible from Python. We are currently integrating the key algorithms
-from Livre for out-of-core volume rendering in Brayns.
-
-\FIXME{teaser gallery}
-
-\subsection{RTNeuron: OpenGL Parallel Rendering}
-
-\subsection{Livre: Out-of-Core Volume Rendering}
-
-Real-time rendering of volumetric data suffers from the curse of
-dimensionality. The data size of regular grids increases cubically as the number
-of dimensions increase. To render such data interactively, out-of-core (OOC)
-storage using multi-resolution algorithms are commonly employed.
-
-Livre is an interactive volume rendering engine available under a permissive
-open source license. Our main contributions are: an state-of-the-art
-implementation of an octree-based level-of-detail selection, a task parallel
-rendering pipeline, a multi-GPU based parallel rendering engine, and an easily
-extensible renderer through the use of plugin data sources. Furthermore we
-propose to create volume representations of different input data using
-on-the-fly, interactive transformations of the data, skipping costly
-preprocessing steps.
-
-Our system brings together state-of-the art algorithms to create a volume
-rendering engine capable of handling extremely high-resolution volumes using a
-high degree of parallelism, both on a single system as well as in a distributed
-cluster. We employ a GPU-based ray casting algorithm to compute the radiance
-absorption of the given volumetric data. The computation is executed per pixel
-on the pixel shader hardware of the GPU.
-
-In our out-of-core data access layer, multi-resolution data is represented as an
-octree data structure. This representation accelerates the selection of the
-proper level-of-detail and to track the status of the LODs (in CPU memory, in
-GPU memory, not loaded). While rendering, view-based LOD selection is performed
-using the on the screen-space-error (SSE)~\cite{guthe2004} technique. According
-to the requested maximum SSE, the selected LODs represent the given volume the
-best on the screen.
-
-The creation of volume bricks, their upload to the GPU and the rendering are
-executed in separate tasks. These tasks run asynchronously, that is, they do not
-block each other. This is particularly important for the render thread, which
-has to react to user input as quickly as possible to reduce input latency, and
-renders at interactive frame rates. The data loader thread can take seconds to
-generate the brick, for example in the field voxelization data source when
-sampling millions of events.
-
-Livre uses a plugin mechanism to access the volume data. In this mechanism, data
-sources are implemented as shared libraries and are loaded on application
-startup based on the URI of the input data. Loading the shared libraries on
-runtime allows to add third-party data loaders to an existing Livre application
-by simply placing them into the shared library directory. Data sources have to
-provide the requested volume bricks, that is, there is no defined file format or
-even requirement to read the input data from a file system. This flexibility of
-the plugin approach lead to novel volume rendering use cases, where volume
-representations are created on the fly from different input data sets.
-
-\FIXME{teaser gallery}
-
 %----------------------------------------------------------------------

### 2017-05-29 paper.tex
Draft conclusion

@@
 
-\section{C++ Messaging and Services}
+\section{Messaging and Service Architecture}
 
@@
 
-Brayns will become the integrative visualizer
-
-Messaging enables surprising new use cases
-
-Tiled display walls are cheap and ready for new ways of working
+We have presented a modular visualization architecture for large data
+visualization over a wide range of use cases, glued together by a modern and
+easy to use messaging infrastructure. This ecosystem allows us to flexibly
+support novel use cases, while pushing novel visualization capabilities,
+providing Blue Brain with a competitive visualization infrastructure. Messaging
+and remote APIs not only surprised us in their versatility and ease of
+integration with other tools, but have still a significant potential for future
+exploration of classical visualization software. Interactive raytracing is the
+rendering algorithm for us, and Brayns is becoming the integration point for our
+domain-specific visual applications and algorithms. Tiled display walls are
+affordable for a large set of institutions, and coupled with our open source
+TIDE software create new ways of truely collaborative work. TIDE, together with
+cheaper visualization hardware, will evolve in the future towards a seamless
+integration of immersive visualization.
 

### 2017-05-30 paper.tex
Review Livre section

@@
 
-Real-time rendering of volumetric data suffers from the curse of
-dimensionality. The data size of regular grids increases cubically as the number
-of dimensions increase. To render such data interactively, out-of-core (OOC)
-storage using multi-resolution algorithms are commonly employed.
-
 Livre is an interactive volume rendering engine available under a permissive
-open source license. Our main contributions are: an state-of-the-art
-implementation of an octree-based level-of-detail selection, a task parallel
-rendering pipeline, a multi-GPU based parallel rendering engine, and an easily
-extensible renderer through the use of plugin data sources. Furthermore we
-propose to create volume representations of different input data using
-on-the-fly, interactive transformations of the data, skipping costly
-preprocessing steps.
-
-Our system brings together state-of-the art algorithms to create a volume
-rendering engine capable of handling extremely high-resolution volumes using a
-high degree of parallelism, both on a single system as well as in a distributed
-cluster. We employ a GPU-based ray casting algorithm to compute the radiance
-absorption of the given volumetric data. The computation is executed per pixel
-on the pixel shader hardware of the GPU.
+open source license. Our main contributions are: a state-of-the-art
+implementation of an octree-based level-of-detail (LOD) selection, a task
+parallel rendering pipeline, a multi-GPU parallel rendering engine, and an
+easily extensible renderer through the use of plugin data sources. Our system
+brings together state-of-the art algorithms to create a volume rendering engine
+capable of handling extremely high-resolution volumes using a high degree of
+parallelism, both on a single system and in a distributed cluster. We employ a
+GPU-based ray casting algorithm to compute the radiance absorption of the given
+volumetric data. The computation is executed per pixel on the pixel shader
+hardware of the GPU.
 
@@
 GPU memory, not loaded). While rendering, view-based LOD selection is performed
-using the on the screen-space-error (SSE)~\cite{guthe2004} technique. According
-to the requested maximum SSE, the selected LODs represent the given volume the
-best on the screen.
+using the screen-space-error (SSE)~\cite{guthe2004} technique. \fig{fLivre}
+shows the rendering of 300 GB volume with an illustration of the selected LOD
+levels.
+
+\begin{figure}[h!t]
+    \includegraphics[width=.45\columnwidth]{images/livre}\hfill
+    \includegraphics[width=.45\columnwidth]{images/livreLOD}
+  \caption{\label{fLivre}Livre rendering a 300GB vasculature dataset (left) and
+    the selected LODs (right)}
+\end{figure}
 
 The creation of volume bricks, their upload to the GPU and the rendering are
-executed in separate tasks. These tasks run asynchronously, that is, they do not
-block each other. This is particularly important for the render thread, which
-has to react to user input as quickly as possible to reduce input latency, and
-renders at interactive frame rates. The data loader thread can take seconds to
-generate the brick, for example in the field voxelization data source when
-sampling millions of events.
-
-Livre uses a plugin mechanism to access the volume data. In this mechanism, data
-sources are implemented as shared libraries and are loaded on application
-startup based on the URI of the input data. Loading the shared libraries on
-runtime allows to add third-party data loaders to an existing Livre application
-by simply placing them into the shared library directory. Data sources have to
-provide the requested volume bricks, that is, there is no defined file format or
-even requirement to read the input data from a file system. This flexibility of
-the plugin approach lead to novel volume rendering use cases, where volume
-representations are created on the fly from different input data sets.
-
-\FIXME{teaser gallery}
+executed in separate tasks. These tasks run asynchronously, that is, rendering
+is decoupled from data loading. Livre uses a plugin mechanism to access the
+volume data, where data sources are implemented as shared libraries and are
+loaded on application startup based on the URI of the input data. Data sources
+only have to provide the requested volume bricks, that is, there is no defined
+file format or even requirement to read the input data from a file system. This
+flexibility of the plugin approach lead to novel volume rendering use cases,
+where volume representations are created on the fly from different input data
+sets, for example from simulation data.
 
@@
 
-We have presented a modular visualization architecture for large data
-visualization over a wide range of use cases, glued together by a modern and
-easy to use messaging infrastructure. This ecosystem allows us to flexibly
-support novel use cases, while pushing novel visualization capabilities,
-providing Blue Brain with a competitive visualization infrastructure. Messaging
-and remote APIs not only surprised us in their versatility and ease of
-integration with other tools, but have still a significant potential for future
-exploration of classical visualization software. Interactive raytracing is the
-rendering algorithm for us, and Brayns is becoming the integration point for our
-domain-specific visual applications and algorithms. Tiled display walls are
-affordable for a large set of institutions, and coupled with our open source
-TIDE software create new ways of truely collaborative work. TIDE, together with
-cheaper visualization hardware, will evolve in the future towards a seamless
-integration of immersive visualization.
+We presented a modular visualization architecture for large data visualization
+over a wide range of use cases, glued together by a modern and easy to use
+messaging infrastructure. This ecosystem allows us to flexibly support novel use
+cases, while pushing novel visualization capabilities, providing Blue Brain with
+a competitive visualization infrastructure. Messaging and remote APIs not only
+surprised us in their versatility and ease of integration with other ecosystems,
+but also have a significant potential for future exploration in classical
+visualization software. Interactive raytracing is the future rendering algorithm
+for us, and Brayns is becoming the integration point for our domain-specific
+visual applications and algorithms. Tiled display walls are affordable for a
+large set of institutions, and coupled with our open source TIDE software create
+new ways of truely collaborative work. TIDE, together with cheaper visualization
+hardware, will evolve in the future towards a seamless integration of immersive
+visualization.
 

### 2017-05-30 paper.tex
Complete Acknowledgements

@@
 Innovation under grant (TIN2010-21289-C02-01/02), the Cajal Blue Brain Project,
-Hasler and HBP.
+Hasler Stiftung Projekt Nr. 12097, and from the European Union’s Horizon 2020
+research and innovation programme under grant agreement No 720270 (HBP SGA1).
 

### 2017-06-02 paper.tex
Minor Livre tweak
Abstract

@@
 \begin{abstract}
-  The Blue Brain project has been investing in high-performance visualization
-  (HPV) to complement its HPC strategy since its inception in 2007. In 2011,
-  this strategy has been accelerated to develop innovative visualization
-  solutions through through increased funding and strategic partnerships with
-  other research institutions.
+  Blue Brain has pushed high-performance visualization (HPV) to complement its
+  HPC strategy since its inception in 2007. In 2011, this strategy has been
+  accelerated to develop innovative visualization solutions through increased
+  funding and strategic partnerships with other research institutions.
 
@@
   visualization applications with novel collaborative display systems. We
-  motivate how our strategy of transforming high-performance visualization
-  engines into services enables a variety of use cases, not only for the
-  integration with high-fidelity displays, but also to build service oriented
-  architectures, to link into web applications and to provide remote services to
-  python applications.
+  motivate how our strategy of transforming visualization engines into services
+  enables a variety of use cases, not only for the integration with
+  high-fidelity displays, but also to build service oriented architectures, to
+  link into web applications and to provide remote services to python
+  applications.
 \end{abstract}
@@
 
-Visualization supports the BBP along all parts of this project to understand and
+Visualization supports the BBP along all parts of the project to understand and
 debug model data, building and simulation algorithms as well as validating and
@@
 using the screen-space-error (SSE)~\cite{guthe2004} technique. \fig{fLivre}
-shows the rendering of 300 GB volume with an illustration of the selected LOD
-levels.
+shows the rendering of a MicroCT dataset with an illustration of the selected
+LOD levels.
 
@@
     \includegraphics[width=.45\columnwidth]{images/livreLOD}
-  \caption{\label{fLivre}Livre rendering a 300GB vasculature dataset (left) and
+  \caption{\label{fLivre}Livre rendering a MicroCT dataset (left) and
     the selected LODs (right)}
@@
   \includegraphics[width=.382\textwidth]{images/ZeroMS}
-  {\caption{\label{fZero}Messaging enables new use cases}}\vspace{-2ex}
+  {\caption{\label{fZero}New messaging use cases}}\vspace{-2ex}
 \end{wrapfigure}

### 2017-06-02 paper.tex
Tide feedback

@@
 
-The Blue Brain Project (BBP) uses simulation-based research to analyze and reverse
-engineer cortical neuron circuits. The simulated models go beyond using detailed
-models of individual neurons or large-scale network models of simplified
-neurons, they model in the order of hundreds of thousands of detailed neurons in
-a fully connected circuit. The project generates a multitude of data for the
-model building and during the simulation of these models. This data ranges from
-terabyte-sized image stacks for data extraction to detailed in-silico circuit
-models of large geometric complexity and terabyte-size simulation reports.
+The Blue Brain Project (BBP) uses simulation-based research to analyze and
+reverse engineer cortical neuron circuits. The simulated models go beyond using
+detailed models of individual neurons or large-scale network models of
+simplified neurons, they model in the order of hundreds of thousands of detailed
+neurons in a fully connected circuit. The project generates a multitude of data
+for the model building and during the simulation of these models. This data
+ranges from terabyte-sized image stacks for data extraction to detailed
+in-silico circuit models of large geometric complexity and terabyte-size
+simulation reports.
 
@@
 
-\subsection{Tiled Multitouch Display Walls}
-
-The core of the Blue Brain visualization infrastructure are multiple
-high-resolution tiled display walls driven by our Tide software~\cite{tide}. All
-walls are equipped with a multitouch user interface and can be remote controlled
-from any web browser. The walls are build using thin-bezel, 55 inch, Full HD LCD
-panels with a hardened glass sheet. We use $4\times 3$ and $3\times 3$
-configurations for a total of 24 and 18 Megapixel resolution, respectively. The
-display size of over five meter diagonally (four meter for $3\times 3$) allows
-team-size collaboration (up to ten people) or project-wide presentations (up to
-a hundred people). \fig{fTDW} shows one wall during a project-wide presentation
-with multiple interactive applications running remotely on the wall.
-
-\begin{figure}[h!t]
-  \includegraphics[width=\columnwidth]{images/tdw}
-  \caption{\label{fTDW}Blue Brain $4\times 3$ tiled display wall}
-\end{figure}
-
-While there is substantial research on tiled display wall software~\cite{Sage,
-Sage2, Omegalib, DK:11, JLMV:06, DisplayCluster}, we found that most solutions
-where not ready production use in a $24\times 7$ unattained environment. For
-this reason, we started with the TACC DisplayCluster open source
-software~\cite{DisplayCluster}, which we incrementally refactored and improved
-to the current TIDE implementation. On the other hand, the hardware has been
-commoditized to make these type of installations affordable to medium-sized
-institutions which allowed us to build the software integration for a reasonable
-startup cost. We have focused on the multitouch user interface, which implements
-a low entry barrier for new users, a unique capability of our solution.
-
-\subsection{OpenDeck}
-
-OpenDeck is our next-generation visualization system, aiming to integrate the
-success of tiled display walls with a seamless transition to fully immersive
-environments. We are currently in the process of installing a system which
-consists of a semi-cylindrical back-projection screen with 41 Megapixel usable
-resolution on a 36 $m^2$ surface (\fig{fOpenDeck}). Like the display walls, it
-is equipped with multitouch capabilities which makes it usable as a monoscopic
-collaboration system from the first day of installation. Unlike tiled display
-walls, it is active stereo capable and equipped with a 3D tracking system for
-immersive rendering. For increased immersion, a lower resolution front
-projection system fills in the floor area. OpenDeck will run TIDA and our
-immersive applications based on Equalizer~\cite{EMP:09} once the system is
-installed.
-
-\begin{figure}[h!t]
-  \includegraphics[width=\columnwidth]{images/opendeck}
-  \caption{\label{fOpenDeck}OpenDeck concept rendering}
-\end{figure}
-
-OpenDeck will provide a unique environment for the research and development of
-new visualization techniques. It will open a set of questions along immersive
-touch user interfaces, transitions and mixing of monoscopic to immersive usage,
-the combination of tracked and touch devices, multi-user immersion, latency for
-remote immersive rendering as well as multi-site collaboration.
-
 \subsection{Tide}
@@
 
-Compared to other solutions~\cite{Sage,Sage2,DisplayCluster}, the development of
-Tide was driven by the need of a stable and easy to use implementation. Tide
-supports three types of content: files (high-resolution images, movies, pdfs),
-built-in applications (web browser, whiteboard) and remote applications using
-the Deflect library (DesktopStreamer, Equalizer-based applications, Brayns).
-The multitouch user interface can handle multiple users manipulating different
-windows and their content simultaneously.
+While there is substantial research on tiled display wall software~\cite{Sage,
+Sage2, Omegalib, DK:11, JLMV:06, DisplayCluster}, we found that most solutions
+were not ready for production use in a $24\times 7$ unattended environment. For
+this reason, we started with the TACC DisplayCluster open source
+software~\cite{DisplayCluster}, which we incrementally refactored and improved
+to the current TIDE implementation. On the other hand, the hardware has been
+commoditized to make these type of installations affordable to medium-sized
+institutions which allowed us to build the software integration for a reasonable
+startup cost. We have focused on the multitouch user interface, which implements
+a low entry barrier for new users, a unique capability of our solution.
+
+Tide supports three types of content: files (high-resolution images, movies,
+pdfs), built-in applications (web browser, whiteboard) and remote applications
+using the Deflect library (DesktopStreamer, Equalizer-based applications,
+Brayns). The multitouch user interface can handle multiple users manipulating
+different windows and their content simultaneously.
+
+\subsection{Tiled Multitouch Display Walls}
+
+The core of the Blue Brain visualization infrastructure are multiple
+high-resolution tiled display walls driven by our Tide software~\cite{tide}. All
+walls are equipped with a multitouch user interface and can be remote controlled
+from any web browser. The walls are built using thin-bezel, 55 inch, Full HD LCD
+panels with a hardened glass sheet. We use $4\times 3$ and $3\times 3$
+configurations for a total of 24 and 18 Megapixel resolution, respectively. The
+display size of over five meter diagonally (four meter for $3\times 3$) allows
+team-size collaboration (up to ten people) or project-wide presentations (up to
+a hundred people). \fig{fTDW} shows one wall during a project-wide presentation
+with multiple interactive applications running remotely on the wall.
+
+\begin{figure}[h!t]
+  \includegraphics[width=\columnwidth]{images/tdw}
+  \caption{\label{fTDW}Blue Brain $4\times 3$ tiled display wall}
+\end{figure}
+
+\subsection{OpenDeck}
+
+OpenDeck is our next-generation visualization system, aiming to integrate the
+success of tiled display walls with a seamless transition to fully immersive
+environments. We are currently in the process of installing a system which
+consists of a semi-cylindrical back-projection screen with 41 Megapixel usable
+resolution on a 36 $m^2$ surface (\fig{fOpenDeck}). Like the display walls, it
+is equipped with multitouch capabilities which makes it usable as a monoscopic
+collaboration system from the first day of installation. Unlike tiled display
+walls, it is active stereo capable and equipped with a 3D tracking system for
+immersive rendering. For increased immersion, a lower resolution front
+projection system fills in the floor area. OpenDeck will run TIDE and our
+immersive applications based on Equalizer~\cite{EMP:09} once the system is
+installed.
+
+\begin{figure}[h!t]
+  \includegraphics[width=\columnwidth]{images/opendeck}
+  \caption{\label{fOpenDeck}OpenDeck concept rendering}
+\end{figure}
+
+OpenDeck will provide a unique environment for the research and development of
+new visualization techniques. It will open a set of questions along immersive
+touch user interfaces, transitions and mixing of monoscopic to immersive usage,
+the combination of tracked and touch devices, multi-user immersion, latency
+reduction for remote immersive rendering as well as multi-site collaboration.
 

### 2017-06-02 paper.tex
Update Authors

@@
 %% Author and Affiliation (multiple authors with multiple affiliations)
-\author{Stefan~Eilemann, et.al.\thanks{\textit{firstname.lastname}@epfl.ch}}
+\author{\small Stefan~Eilemann, Marwan~Abdellah, Nicolas~Antille, Ahmet~Bilgili, Grigory~Chevtchenko, Raphael~Dumusc, Cyrille~Favreau, Juan~Hernando, Daniel~Nachbaur, Pawel~Podhajski, Jafet~Villafranca\thanks{\textit{firstname.lastname}@epfl.ch}}
 \institute{Blue Brain Project, Ecole Polytechnique Federale de Lausanne}
@@
 Tide (Tiled Interactive Display Environment) is the software driving the Blue
-Brain tiled display walls and OpenDeck. It provides multi-window, multi-user
-touch interaction on large surfaces --- think of a giant collaborative
-wall-mounted tablet. Tide is a distributed application that can run on multiple
-machines to power display walls or projection systems of any size. Its user
-interface is designed to offer an intuitive experience on touch walls. It works
-just as well on non touch-capable installations by using its web interface from
-any web browser. \fig{fTDW} shows Tide on a $4\times 3$ display wall and
-\fig{fTideWeb} shows the Tide web interface in a browser.
+Brain tiled display walls and the future OpenDeck. It provides multi-window,
+multi-user touch interaction on large surfaces --- think of a giant
+collaborative wall-mounted tablet. Tide is a distributed application that can
+run on multiple machines to power display walls or projection systems of any
+size. Its user interface is designed to offer an intuitive experience on touch
+walls. It works just as well on non touch-capable installations by using its web
+interface from any web browser. \fig{fTDW} shows Tide on a $4\times 3$ display
+wall and \fig{fTideWeb} shows the Tide web interface in a browser.
 

### 2017-06-02 paper.tex
RTNeuron description and review up to that point.

@@
   high-fidelity displays, but also to build service oriented architectures, to
-  link into web applications and to provide remote services to python
+  link into web applications and to provide remote services to Python
   applications.
@@
 standard rendering engines (rasterization, out-of-core volume rendering,
-interactive raytracing), and decoupled, light-weight applications using remote
-these components. In the following we will present these components along with a
-few use cases.
+interactive raytracing), and decoupled, light-weight applications using
+these components remotely. In the following we will present these components
+along with a few use cases.
 
@@
 visualization capabilities. They are the evolution of existing visualization
-systems, enabling higher resolution, immersion and team collaboration. Compared
+systems, enabling high resolution, immersion and team collaboration. Compared
 to current single-user or single-presenter systems, collaborative display
@@
 exploration.
-
-\subsection{Tide}
-
-Tide (Tiled Interactive Display Environment) is the software driving the Blue
-Brain tiled display walls and the future OpenDeck. It provides multi-window,
-multi-user touch interaction on large surfaces --- think of a giant
-collaborative wall-mounted tablet. Tide is a distributed application that can
-run on multiple machines to power display walls or projection systems of any
-size. Its user interface is designed to offer an intuitive experience on touch
-walls. It works just as well on non touch-capable installations by using its web
-interface from any web browser. \fig{fTDW} shows Tide on a $4\times 3$ display
-wall and \fig{fTideWeb} shows the Tide web interface in a browser.
-
-\begin{figure}[h!t]
-  \includegraphics[width=\columnwidth]{images/tideweb}
-  \caption{\label{fTideWeb}Tide web interface}
-\end{figure}
-
-While there is substantial research on tiled display wall software~\cite{Sage,
-Sage2, Omegalib, DK:11, JLMV:06, DisplayCluster}, we found that most solutions
-were not ready for production use in a $24\times 7$ unattended environment. For
-this reason, we started with the TACC DisplayCluster open source
-software~\cite{DisplayCluster}, which we incrementally refactored and improved
-to the current TIDE implementation. On the other hand, the hardware has been
-commoditized to make these type of installations affordable to medium-sized
-institutions which allowed us to build the software integration for a reasonable
-startup cost. We have focused on the multitouch user interface, which implements
-a low entry barrier for new users, a unique capability of our solution.
-
-Tide supports three types of content: files (high-resolution images, movies,
-pdfs), built-in applications (web browser, whiteboard) and remote applications
-using the Deflect library (DesktopStreamer, Equalizer-based applications,
-Brayns). The multitouch user interface can handle multiple users manipulating
-different windows and their content simultaneously.
 
@@
 \end{figure}
+
+\subsection{Tide}
+
+Tide (Tiled Interactive Display Environment) is the software driving the Blue
+Brain tiled display walls. It provides multi-window, multi-user touch
+interaction on large surfaces --- think of a giant collaborative wall-mounted
+tablet. Tide is a distributed application that can run on multiple machines to
+power display walls or projection systems of any size. Its user interface is
+designed to offer an intuitive experience on touch walls. It works just as
+well on non touch-capable installations by using its web interface from any
+web browser. \fig{fTDW} shows Tide on a $4\times 3$ display wall and
+\fig{fTideWeb} shows the Tide web interface in a browser.
+
+\begin{figure}[h!t]
+  \includegraphics[width=\columnwidth]{images/tideweb}
+  \caption{\label{fTideWeb}Tide web interface}
+\end{figure}
+
+While there is substantial research on tiled display wall software~\cite{Sage,
+Sage2, Omegalib, DK:11, JLMV:06, DisplayCluster}, we found that most solutions
+were not ready for production use in a $24\times 7$ unattended environment. For
+this reason, we started with the TACC DisplayCluster open source
+software~\cite{DisplayCluster}, which we incrementally refactored and improved
+to the current TIDE implementation. On the other hand, the hardware has been
+commoditized to make these type of installations affordable to medium-sized
+institutions which allowed us to build the software integration for a reasonable
+startup cost. % I don't understand very well what do you try to mean with this last sentence.
+We have focused on the multitouch user interface, which implements
+a low entry barrier for new users, a unique capability of our solution.
+
+Tide supports three types of content: files (high-resolution images, movies,
+pdfs), built-in applications (web browser, whiteboard) and remote applications
+using the Deflect library (DesktopStreamer, Equalizer-based applications,
+Brayns). The multitouch user interface can handle multiple users manipulating
+different windows and their content simultaneously.
+
+\subsection{Deflect}
+
+Deflect is the client library for Tide. It provides an API for pixel streaming
+to Tide and for receiving events from Tide. The pixel streaming allows
+synchronized parallel streaming from parallel rendering application as well as
+monoscopic and stereoscopic streams. Various events allow the application to
+react to multi-touch input from the wall.
+
+Deflect is integrated into the Equalizer parallel rendering
+framework~\cite{EMP:09}, enabling transparent usage of Equalizer applications on
+Tide walls. Furthermore, the DesktopStreamer application mirrors the desktop of
+other machines onto a wall window and allows interaction with the remote
+desktop. Other rendering applications, such as our interactive raytracing
+engine Brayns~\cite{brayns} are easily integrated with Deflect and Tide.
 
@@
 
-\subsection{Deflect}
-
-Deflect is the client library for Tide. It provides an API for pixel streaming
-to Tide and for receiving events from Tide. The pixel streaming allows
-synchronized parallel streaming from parallel rendering application as well as
-monoscopic and stereoscopic streams. Various events allow the application to
-react to multi-touch input from the wall.
-
-Deflect is integrated into the Equalizer parallel rendering
-framework~\cite{EMP:09}, enabling transparent usage of Equalizer applications on
-Tide walls. Furthermore, the DesktopStreamer application mirrors the desktop of
-other machines onto a wall window and allows interaction with the remote
-desktop. Other rendering applications, such as our interactive raytracing
-engine Brayns~\cite{brayns} are easily integrated with Deflect and Tide.
-
 \section{Rendering Applications}
@@
 is for rendering at very high frame rates, needed for example for immersive
-visualization.
+visualization, or very high resolutions.
 
@@
 Livre for out-of-core volume rendering in Brayns, which will also facilitate
-mixing polygonal with volume data in a single scene.
+mixing polygonal with large volumetric data in a single scene.
 
@@
 \subsection{RTNeuron: OpenGL Parallel Rendering}
+
+RTNeuron is the oldest of our interactive visualization tools. Originally
+conceived as a standalone application for visualizing simulation results, it
+has evolved into a rasterization-based rendering engine library implemented in
+C++ with a Python wrapping. The power of RTNeuron lies in a domain specific API
+designed for the visualization of detailed neuronal circuit models, this API
+allows for building custom applications tailored for specific use cases.
+It provides features to visualize static circuit data and some simulation
+results. Static data visualization includes different visual representations for
+neurons, synapses, selective pruning of neuronal trees, clipping
+planes... The simulation results that can be displayed are the spiking
+activity of the neurons and scalar variables from the cable models mapped to
+the neuron surface. The color maps for displaying these data are highly
+configurable, being posible to apply different color maps to different cell
+subsets.
+
+Neurons can be displayed with different levels of detail ranging from simple
+spheres, cylinder-like geometric models and detailed polygonal
+meshes. Advanced rendering techniques included are several types of parallel
+rendering algorithms (such as sort-first and sort-last) and several algorithms
+for transparency rendering that enable efficient and correct rendering of
+scenes with great geometrical complexity and high depth
+complexity. Transparency is particularly suitable for masking and highlighting
+features of interest on the circuit. Ongoing work is a more scalable parallel
+rendering algorithm specially suited for transparency. Tiled rendering is also
+possible, which allows very high resolution renderings suitable for printing
+at sizes larger than A0.
+
+The core engine is implemented in C++ using Equalizer and OpenSceneGraph and
+also leverages part of our messaging framework to allow coupling with other
+applications. The 3D view can be embedded inside Qt applications, in particular
+in Python with PyQt and QML for overlaying GUI elements. Several use-case
+specific applications have been built this way.
 
@@
 %----------------------------------------------------------------------
+
 This publication was supported by the Blue Brain Project (BBP), the Swiss
@@
 
+We would also like to thank the people from GMRV at the Rey Juan Carlos
+University (URJC) for their collaboration under the Cajal Blue Brain and HBP
+projects.
+
 \bibliographystyle{abbrv}

### 2017-06-05 paper.tex
Update pdf

@@
 %% Author and Affiliation (multiple authors with multiple affiliations)
-\author{\small Stefan~Eilemann, Marwan~Abdellah, Nicolas~Antille, Ahmet~Bilgili, Grigory~Chevtchenko, Raphael~Dumusc, Cyrille~Favreau, Juan~Hernando, Daniel~Nachbaur, Pawel~Podhajski, Jafet~Villafranca\thanks{\textit{firstname.lastname}@epfl.ch}}
+\author{Stefan~Eilemann, Marwan~Abdellah, Nicolas~Antille, Ahmet~Bilgili, Grigory~Chevtchenko, Raphael~Dumusc, Cyrille~Favreau, Juan~Hernando, Daniel~Nachbaur, Pawel~Podhajski, Jafet~Villafranca, Felix Sch\"urmann\thanks{\textit{firstname.lastname}@epfl.ch}}
 \institute{Blue Brain Project, Ecole Polytechnique Federale de Lausanne}
@@
 Hasler Stiftung Projekt Nr. 12097, and from the European Union’s Horizon 2020
-research and innovation programme under grant agreement No 720270 (HBP SGA1).
-
-We would also like to thank the people from GMRV at the Rey Juan Carlos
-University (URJC) for their collaboration under the Cajal Blue Brain and HBP
-projects.
+research and innovation programme under grant agreement No 720270 (HBP SGA1). We
+would also like to thank the people from GMRV at the Rey Juan Carlos University
+(URJC) for their collaboration under the Cajal Blue Brain and HBP projects.
 

### 2017-06-06 paper.tex
Brayns gallery

@@
 memory size. Last, but not least, advanced rendering algorithms such as
-shadowing, reflections and global illuminations are significantly easier to
-implement in a raytracer. The only area where rasterization provides a benefit
-is for rendering at very high frame rates, needed for example for immersive
-visualization, or very high resolutions.
+shadowing, reflections and global illumination (\fig{fBrayns}, left) are
+significantly easier to implement in a raytracer. The only area where
+rasterization provides a benefit is for rendering at very high frame rates,
+needed for example for immersive visualization, or very high resolutions.
+
+\begin{figure}[ht]\center
+  \includegraphics[height=5cm]{images/Brayns003}\hfil
+  \includegraphics[height=5cm]{images/Brayns001}
+  \caption{\label{fBrayns}Interactive raytracing for neuroscience}
+\end{figure}
 
@@
 Livre for out-of-core volume rendering in Brayns, which will also facilitate
-mixing polygonal with large volumetric data in a single scene.
-
-\FIXME{teaser gallery}
+mixing polygonal with large volumetric data in a single scene (\fig{fBrayns},
+right).
 

### 2017-06-06 paper.tex
Tweak RTNeuron section

@@
 % correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor Zero-Buf}
+\hyphenation{op-tical net-works semi-conduc-tor Zero-Buf ras-te-ri-za-tion}
 
@@
 
-RTNeuron is the oldest of our interactive visualization tools. Originally
-conceived as a standalone application for visualizing simulation results, it
-has evolved into a rasterization-based rendering engine library implemented in
-C++ with a Python wrapping. The power of RTNeuron lies in a domain specific API
-designed for the visualization of detailed neuronal circuit models, this API
-allows for building custom applications tailored for specific use cases.
-It provides features to visualize static circuit data and some simulation
-results. Static data visualization includes different visual representations for
-neurons, synapses, selective pruning of neuronal trees, clipping
-planes... The simulation results that can be displayed are the spiking
-activity of the neurons and scalar variables from the cable models mapped to
-the neuron surface. The color maps for displaying these data are highly
-configurable, being posible to apply different color maps to different cell
+RTNeuron\cite{Hernando13} is the oldest of our interactive visualization tools.
+Originally conceived as a standalone application for visualizing simulation
+results, it has evolved into a rasterization-based rendering engine library
+implemented in C++ with a Python wrapping. The power of RTNeuron lies in a
+domain specific API designed for the visualization of detailed neuronal circuit
+models, this API allows for building custom applications tailored for specific
+use cases. It provides features to visualize static circuit data and some
+simulation results. Static data visualization includes different visual
+representations for neurons, synapses, selective pruning of neuronal trees,
+clipping planes and others. The simulation results that can be displayed are the
+spiking activity of the neurons and scalar variables from the cable models
+mapped to the neuron surface. The color maps for displaying these data are
+highly configurable, allowing to apply different color maps to different cell
 subsets.
@@
 The core engine is implemented in C++ using Equalizer and OpenSceneGraph and
-also leverages part of our messaging framework to allow coupling with other
+leverages part of our messaging framework to allow coupling with other
 applications. The 3D view can be embedded inside Qt applications, in particular
-in Python with PyQt and QML for overlaying GUI elements. Several use-case
+in Python with PyQt and QML for overlaying GUI elements. Several use case
 specific applications have been built this way.

### 2017-06-20 paper.tex
RTNeuron images.

@@
 
-RTNeuron\cite{Hernando13} is the oldest of our interactive visualization tools.
+RTNeuron\cite{HBBES:13} is the oldest of our interactive visualization tools.
 Originally conceived as a standalone application for visualizing simulation
@@
 possible, which allows very high resolution renderings suitable for printing
-at sizes larger than A0.
+at sizes larger than A0 (The left side of \fig{fRTNeuron} has been rendered at
+36640$\times$26000).
 
@@
 leverages part of our messaging framework to allow coupling with other
-applications. The 3D view can be embedded inside Qt applications, in particular
-in Python with PyQt and QML for overlaying GUI elements. Several use case
-specific applications have been built this way.
+applications. The 3D view can be embedded inside Qt applications
+(\fig{fRTNeuron}, right side), in particular in Python with PyQt and QML for
+overlaying GUI elements. Several use case specific applications have been built
+this way.
+
+\begin{figure}[ht]\center
+  \includegraphics[width=.45\columnwidth]{images/rtneuron1.png}\hfil
+  \includegraphics[width=.45\columnwidth]{images/rtneuron2.png}
+  \caption{\label{fRTNeuron}Left, membrame potential of a fraction of the neurons in a simulation of neocortical circuit.
+Right, a circuit visualization application showing a portion of a hippocampus model.}
+\end{figure}
 

### 2017-06-20 paper.tex
Layout and wording changes.

--- 2017-06-06 paper.tex
@@
 possible, which allows very high resolution renderings suitable for printing
-at sizes larger than A0.
+at sizes larger than A0 (The original of \fig{fRTNeuron} left is a 36640$\times$26000 pixel image).
 
@@
 leverages part of our messaging framework to allow coupling with other
-applications. The 3D view can be embedded inside Qt applications, in particular
-in Python with PyQt and QML for overlaying GUI elements. Several use case
-specific applications have been built this way.
+applications. The 3D view can be embedded inside Qt applications
+(\fig{fRTNeuron}, right), in particular in Python with PyQt and QML for
+overlaying GUI elements. Several use case specific applications have been built
+this way.
+
+\begin{figure}[h!t]\center
+  \includegraphics[width=.45\columnwidth]{images/rtneuron1.png}\hfil
+  \includegraphics[width=.45\columnwidth]{images/rtneuron2.png}
+  \caption{\label{fRTNeuron}RTNeuron examples: membrame potential of a fraction of the neurons in a simulation (left) and a circuit visualization application showing a portion of a hippocampus model (right).}
+\end{figure}
 
@@
 
-\begin{figure}[h!t]
+\begin{figure}[t]
     \includegraphics[width=.45\columnwidth]{images/livre}\hfill
@@
   \includegraphics[width=.382\textwidth]{images/ZeroMS}
-  {\caption{\label{fZero}New messaging use cases}}\vspace{-2ex}
+  {\caption{\label{fZero}New messaging use cases}}\vspace{-6em}
 \end{wrapfigure}

### 2017-06-20 paper.tex
Minor tweaks

@@
 Neurons can be displayed with different levels of detail ranging from simple
-spheres, cylinder-like geometric models and detailed polygonal
-meshes. Advanced rendering techniques included are several types of parallel
-rendering algorithms (such as sort-first and sort-last) and several algorithms
-for transparency rendering that enable efficient and correct rendering of
-scenes with great geometrical complexity and high depth
-complexity. Transparency is particularly suitable for masking and highlighting
-features of interest on the circuit. Ongoing work is a more scalable parallel
-rendering algorithm specially suited for transparency. Tiled rendering is also
-possible, which allows very high resolution renderings suitable for printing
-at sizes larger than A0 (The original of \fig{fRTNeuron} left is a 36640$\times$26000 pixel image).
+spheres, cylinder-like geometric models and detailed polygonal meshes. Advanced
+rendering techniques included are several types of parallel rendering algorithms
+(such as sort-first and sort-last) and several algorithms for transparency
+rendering that enable efficient and correct rendering of scenes with great
+geometrical complexity and high depth complexity. Transparency is particularly
+suitable for masking and highlighting features of interest on the circuit.
+Ongoing work is a more scalable parallel rendering algorithm specially suited
+for transparency. Tiled rendering is also possible, which allows very high
+resolution renderings suitable for printing at sizes larger than A0 (the
+original of \fig{fRTNeuron} left is a 36640$\times$26000 pixel image).
 
@@
 \begin{figure}[h!t]\center
-  \includegraphics[width=.45\columnwidth]{images/rtneuron1.png}\hfil
-  \includegraphics[width=.45\columnwidth]{images/rtneuron2.png}
-  \caption{\label{fRTNeuron}RTNeuron examples: membrame potential of a fraction of the neurons in a simulation (left) and a circuit visualization application showing a portion of a hippocampus model (right).}
+  \includegraphics[width=.49\columnwidth]{images/rtneuron1.png}\hfil
+  \includegraphics[width=.49\columnwidth]{images/rtneuron2.png}
+  \caption{\label{fRTNeuron}RTNeuron rendering the simulated membrame potential of a fraction of the simulation~(left) and a circuit visualization application for hippocampus model validation~(right)}
 \end{figure}

### 2017-06-20 paper.tex
Minor tweaks

@@
 \begin{figure}[t]
-    \includegraphics[width=.45\columnwidth]{images/livre}\hfill
-    \includegraphics[width=.45\columnwidth]{images/livreLOD}
+    \includegraphics[width=.49\columnwidth]{images/livre}\hfill
+    \includegraphics[width=.49\columnwidth]{images/livreLOD}
   \caption{\label{fLivre}Livre rendering a MicroCT dataset (left) and
@@
 
+\clearpage
 \section{Messaging and Service Architecture}
@@
   \includegraphics[width=.382\textwidth]{images/ZeroMS}
-  {\caption{\label{fZero}New messaging use cases}}\vspace{-6em}
+  {\caption{\label{fZero}Messaging-enabled use cases}}\vspace{-1em}
 \end{wrapfigure}
@@
 {
-    subscriber.receive( 0 /*ms, poll*/ );
-    updateCamera( camera );
+    if( updateCamera( camera )) // had user event
         publisher.publish( camera );
+    else // poll subscription
+        subscriber.receive( 0 /*ms*/ );
     renderFrame( camera );
